Missing data: Current practice in football research and recommendations for improvement

### 2020-10-22 paper/paper.Rmd
fix from work session

```{r read-data, message = FALSE}
gs_football <- readr::read_rds(here::here("data",
                           "gs-football.rds"))
```

```{r mismatch}
# Now we bring the third and final review 
gs_mismatches <- readr::read_csv(here::here("data",
"missing-data-mismatches.csv"))

gs_mismatch_pivot <- gs_mismatches %>% 
  pivot_longer(cols = c(reviewer_1,reviewer_2, third_check),
               names_to = "reviewer",
               values_to = "decision") %>% 
  pivot_wider(names_from = variable,
              values_from = decision)
```

@@

Affiliations:

@@

Nicholas J Tierney
Monash University, Department of Econometrics and Business Statistics,
Melbourne, Australia.
Address:
Email:

**Data availability**
The code to reproduce the results and plots in the paper can be accessed at: **Nick to host?**

@@

@@

Missing data are often unavoidable. The reason values go missing, along
with decisions made of how missing data are handled (deleted or
imputed), can have a profound effect on the validity, accuracy and usability 
of a study's results. In this article, we aimed to: estimate
the proportion of studies in football research that included a missing
data statement, highlight several practices to avoid in relation to
missing data, and provide recommendations for exploring, visualising and
reporting missingness. Football related articles, published in 2019 were
studied. A survey of 136 articles, sampled at random, was conducted to
determine whether a missing data statement was included. As expected, the
proportion of studies in football research that included a missing data
statement was low, at only X% (95% confidence interval; X% to X%); suggesting that 
missingness is seldom considered by researchers. It is suspected that this result 
is consistent with practice in the wider sports science literature. We recommend 
researchers describe the number and percentage of missing
values, including when there are no missing values. Exploratory analysis
should be conducted to explore missing values, and visualisations
describing missingness overall should be provided in the paper, or at
least supplementary materials. Missing values should almost always be
imputed, and imputation methods should be explored to ensure they are
appropriately representative.

@@

Our intention was to survey, at random, 10% of the total articles found (after 
title and abstract screening). We were primarily interested in
whether missing data was reported or acknowledged; and if not, whether
a dataset was shared. Where applicable, information relating to how missing
values were treated was also extracted (see Supplement 2). Two authors
independently extracted the data (`r md_reported` agreement). Results are reported
as the proportion and 95% confidence interval. (NOTE: need to get the final third reviewer part in)

@@

```{r}
# NOTE: update with the final third reviewer
# need to update the gs_football
library(binom)
# !!comment to explain NAs
successes <- sum(gs_football$any_missing_data_statement == 1, na.rm = TRUE)
trials <- sum(!is.na(gs_football$any_missing_data_statement))
missing_binom_conf <- binom.confint(x = successes, 
              n = trials, 
              conf.level = 0.95, 
              methods = "exact")

n_screened <- n_distinct(gs_football$key)
miss_reported_pct <- percent(missing_binom_conf$mean, accuracy = 0.1)
miss_reported_pct_lower <- percent(missing_binom_conf$lower, accuracy = 0.1)
miss_reported_pct_upper <- percent(missing_binom_conf$upper, accuracy = 0.1)

n_shared <- sum(gs_football$was_data_shared == 1)
```

```{r}
# now of the articles that don't mention missingness
gs_football_no_mention_sharing <- gs_football %>% 
  filter(any_missing_data_statement == 0)

n_no_mention_missings <- n_distinct(gs_football_no_mention_sharing$key)

no_mention_successes <- sum(gs_football_no_mention_sharing$was_data_shared == 1, 
                            na.rm = TRUE)
no_mention_trials <- sum(!is.na(gs_football_no_mention_sharing$was_data_shared))
n_mention_missing_binom_conf <- binom.confint(x = no_mention_successes, 
                                              n = no_mention_trials, 
                                              conf.level = 0.95, 
                                              methods = "exact")

percent_binom_conf <- function(binom_conf){
  mean_pct <- percent(n_mention_missing_binom_conf$mean, accuracy = 0.1)
lower_pct <- percent(n_mention_missing_binom_conf$lower, accuracy = 0.1)
upper_pct <- percent(n_mention_missing_binom_conf$upper, accuracy = 0.1)

return(c(mean_pct = mean_pct,
         lower_pct = lower_pct,
         upper_pct = upper_pct))
}

percent_binom_conf(n_mention_missing_binom_conf)

Of the 199 articles screened, `r n_screened` met the inclusion criteria, representing
\~10% of the articles found (see Supplement 2, Figure 1). The proportion of
articles that reported missing data was `r miss_reported_pct` (95% confidence interval;
`r miss_reported_pct_lower` to `r miss_reported_pct_upper`) or `r successes`/`r n_screened`.

Of the articles that did not mention missing data `n_no_mention_missings` / `r n_screened`, 7.1% (3.3% to 13.1%) shared their dataset, making
it difficult to determine whether there was any missingness. While the
absence of a missing data statement does not necessarily mean missing
data are not being reported, it does suggest that this aspect of the
analysis process receives little consideration--especially in contrast
to many of the papers that included (potentially meaningless) tests of
normality (e.g., Kruskall-Wallis test).

@@

Our survey of 136 articles published on football related topics found
that only about one in 20 papers reported missing data. It is reasonable
to assume that this result is representative of current practice in the wider sports science literature. 
The low proportion of papers reporting missingness could be explained by 
a lack of awareness and education on missing data [@Sainani2020call], 
similar to other areas of sciences (refs-education and
ecology examples)(2)**references needed**. Below we discuss several practices to avoid in
relation to missingness, and provide recommendations for exploring,
visualising and reporting missing data.

@@

```{r}
data_plot_clean <- data_plot %>% 
  mutate(team = recode(team, 
                        "Gold Coast" = "GC",
                        "Greater Western Sydney" = "GWS",
                        "Fremantle" = "FRE",
                        "Collingwood" = "COL",
                        "Port Adelaide" = "PA",
                        "Western Bulldogs" = "WB",
                        "Richmond" = "RIC",
                        "Brisbane Lions" = "BL",
                        "Geelong" = "GEE",
                        "North Melbourne" = "NM",
                        "Adelaide" = "ADE",
                        "Melbourne" = "MEL",
                        "Sydney" = "SYD",
                        "St Kilda" = "STK",
                        "Essendon" = "ESS",
                        "West Coast" = "WC",
                        "Carlton" = "CAR",
                        "Hawthorn" = "HAW")) %>% 
  as_tibble() %>% 
  rename(sprint_dist = sprint_distance)

>>>>

```{r read-data, message = FALSE, include = FALSE}
gs_football <- readr::read_rds(here::here("data",
                           "gs-football.rds"))
```

```{r mismatch, message = FALSE, include = FALSE}
# Now we bring the third and final review 
gs_mismatches <- readr::read_csv(here::here("data",
"missing-data-mismatches.csv"))

gs_mismatch_pivot <- gs_mismatches %>% 
  pivot_longer(cols = c(reviewer_1,reviewer_2, third_check),
               names_to = "reviewer",
               values_to = "decision") %>% 
  pivot_wider(names_from = variable,
              values_from = decision)

gs_mismatch_pivot %>% 
  slice(1:3)

gs_football
```

```{r finalise-matches}
gs_football_final_matches <- gs_football %>% 
  pivot_longer(cols = any_missing_data_statement:was_data_shared,
               names_to = "variable",
               values_to = "value") %>% 
  pivot_wider(names_from = reviewer) %>% 
  left_join({ 
      select(gs_mismatches, 
             -reviewer_1,
             -reviewer_2)
    }, 
            by = c("variable", "title")) %>% 
  # double check that we get our data back
  # filter(reviewer_1 != reviewer_2)
  mutate(third_check = as.integer(third_check),
         final_check = case_when(
    is.na(third_check) ~ reviewer_1,
    reviewer_1 != reviewer_2 ~ third_check
  ))  %>%  select(-c(reviewer_1, 
            reviewer_2,
            third_check,
            notes)) %>% 
  pivot_wider(names_from = variable,
              values_from = final_check)
  # filter(reviewer_1 != reviewer_2)

@@

Twitter: nj_tierney

Affiliations: 0000-0003-1460-8722

@@

Nicholas J Tierney
Monash University, Department of Econometrics and Business Statistics,
Melbourne, Australia.
Address: 
Email: nicholas.tierney@monash.edu

**Data availability**

The code to reproduce the results and plots in the paper can be accessed at the following github repository:

@@

```{r fun-pct-binom-conf}
percent_binom_conf <- function(binom_conf){
  mean_pct <- percent(binom_conf$mean, accuracy = 0.1)
lower_pct <- percent(binom_conf$lower, accuracy = 0.1)
upper_pct <- percent(binom_conf$upper, accuracy = 0.1)

return(data.frame(mean_pct = mean_pct,
                  lower_pct = lower_pct,
                  upper_pct = upper_pct))
}

```{r binom-ci-fun}
# need to update the gs_football
library(binom)

football_binom_confint <- function(data, var){

  var_vec <- data %>% pull({{ var }})
  successes <- sum(var_vec, na.rm = TRUE)
  trials <- sum(!is.na(var_vec))

  binom.confint(x = successes, 
              n = trials, 
              conf.level = 0.95, 
              methods = "exact")
}
```

```{r binom-ci-reporting}
n_screened <- nrow(gs_football_final_matches)
n_shared <- sum(gs_football_final_matches$was_data_shared == 1)

binom_ci_miss_reported <- football_binom_confint(gs_football_final_matches, 
                                                 any_missing_data_statement) %>% 
  percent_binom_conf()

# now of the articles that don't mention missingness
gs_football_final_matches_dont_report <- gs_football_final_matches %>% 
  filter(any_missing_data_statement == 0)

n_no_mention_missings <- nrow(gs_football_final_matches_dont_report)

binom_ci_data_shared <- gs_football_final_matches_dont_report %>% 
  football_binom_confint(var = was_data_shared) %>% 
  percent_binom_conf()

@@

Missing data are often unavoidable. The reason values go missing, along
with decisions made of how missing data are handled (deleted or
imputed), can have a profound effect on the validity, accuracy and usability 
of a study's results. In this article, we aimed to: estimate
the proportion of studies in football research that included a missing
data statement, highlight several practices to avoid in relation to
missing data, and provide recommendations for exploring, visualising and
reporting missingness. Football related articles, published in 2019 were
studied. A survey of `r n_screened` articles, sampled at random, was conducted to
determine whether a missing data statement was included. As expected, the
proportion of studies in football research that included a missing data
statement was low, at only `r binom_ci_miss_reported$mean_pct`(95% CI; `r binom_ci_miss_reported$lower_pct` to `r binom_ci_miss_reported$upper_pct`); suggesting that 
missingness is seldom considered by researchers. It is suspected that this result 
is consistent with practice in the wider sports science literature. We recommend 
researchers describe the number and percentage of missing
values, including when there are no missing values. Exploratory analysis
should be conducted to explore missing values, and visualisations
describing missingness overall should be provided in the paper, or at
least supplementary materials. Missing values should almost always be
imputed, and imputation methods should be explored to ensure they are
appropriately representative.

@@

Our intention was to survey, at random, 10% of the total articles found (after 
title and abstract screening). We were primarily interested in
whether missing data was reported or acknowledged; and if not, whether
a dataset was shared. Where applicable, information relating to how missing
values were treated was also extracted (see Supplement 2). Two authors
independently extracted the data (`r md_reported` agreement). Disagreements were resolved by author DNB. Results are reported as the proportion and 95% confidence interval (CI).

@@

Of the 199 articles screened, `r n_screened` met the inclusion criteria, representing
\~10% of the articles found (see Supplement 2, Figure 1). The proportion of
articles that reported missing data was `r binom_ci_miss_reported$mean_pct`(95% CI; `r binom_ci_miss_reported$lower_pct` to `r binom_ci_miss_reported$upper_pct`).

Of the articles that did not mention missing data `r n_no_mention_missings` / `r n_screened`, `r binom_ci_data_shared$mean_pct`(95% CI; `r binom_ci_data_shared$lower_pct` to `r binom_ci_data_shared$upper_pct`). shared their dataset, making
it difficult to determine whether there was any missingness. While the
absence of a missing data statement does not necessarily mean missing
data are not being reported, it does suggest that this aspect of the
analysis process receives little consideration--especially in contrast
to many of the papers that included (potentially meaningless) tests of
normality (e.g., Kruskall-Wallis test).

@@

Our survey of `r n_screened` articles published on football related topics found
that only one in 13 papers (`r binom_ci_miss_reported$mean_pct`) reported missing data. It is reasonable
to assume that this result is representative of current practice in the wider sports science literature. 
The low proportion of papers reporting missingness could be explained by 
a lack of awareness and education on missing data [@Sainani2020call], 
similar to other areas of sciences (refs-education and
ecology examples)(2)**references needed**. Below we discuss several practices to avoid in
relation to missingness, and provide recommendations for exploring,
visualising and reporting missing data.

@@

```{r rename-afl-data}
data_plot_clean <- data_plot %>% 
  mutate(team = recode(team, 
                        "Gold Coast" = "GC",
                        "Greater Western Sydney" = "GWS",
                        "Fremantle" = "FRE",
                        "Collingwood" = "COL",
                        "Port Adelaide" = "PA",
                        "Western Bulldogs" = "WB",
                        "Richmond" = "RIC",
                        "Brisbane Lions" = "BL",
                        "Geelong" = "GEE",
                        "North Melbourne" = "NM",
                        "Adelaide" = "ADE",
                        "Melbourne" = "MEL",
                        "Sydney" = "SYD",
                        "St Kilda" = "STK",
                        "Essendon" = "ESS",
                        "West Coast" = "WC",
                        "Carlton" = "CAR",
                        "Hawthorn" = "HAW")) %>% 
  as_tibble() %>% 
  rename(sprint_dist = sprint_distance)
